---
title: "Causal Inference: <br> *The Mixtape*"
subtitle: "<it>Regression Discontinuity</it>"
output: 
  learnr::tutorial:
    css: css/style.css
    highlight: "kate"
runtime: shiny_prerendered
---

## Welcome

This is material for the **Regression Discontinuity** chapter in Scott Cunningham's book, [Causal Inference: The Mixtape.](https://mixtape.scunning.com/)

### Packages needed

The first thing you need to do is install a few packages to make sure everything runs:

```{r, eval = FALSE}
install.packages("tidyverse")
install.packages("cli")
install.packages("haven")
install.packages("rmarkdown")
install.packages("learnr")
install.packages("haven")
install.packages("stargazer")

# This Chapter only
install.packages("texreg")
install.packages("rdrobust")
install.packages("rdd")
install.packages("rddensity")
```

### Load

```{r load, echo=FALSE, warning=FALSE, message=FALSE}
library(learnr)
library(haven)
library(tidyverse)
library(estimatr)
library(stargazer)
library(texreg)

# This Chapter only
library(stats)
library(rdrobust)
library(rddensity)
library(rdd)

# 10 minute code time limit
options(tutorial.exercise.timelimit = 600)

# read_data function
read_data <- function(df) {
  full_path <- paste0("https://raw.github.com/scunning1975/mixtape/master/", df)
  return(haven::read_dta(full_path))
}
```


## RDD Simulation

Generate a simple Regression Discontinuity, before and after the treatment is given. 

```{r rdd_simulate1_2, exercise=TRUE, echo=FALSE}

# simulate the data
dat <- tibble(
    x = rnorm(1000, 50, 25)
  ) %>%
  mutate(
    x = if_else(x < 0, 0, x)
  ) %>%
  filter(x < 100)

# cutoff at x = 50
dat <- dat %>% 
  mutate(
    D  = if_else(x > 50, 1, 0),
    y1 = 25 + 0 * D + 1.5 * x + rnorm(n(), 0, 20)
  )

cli::cli_h1("Counterfactual Potential Outcomes")
ggplot(aes(x, y1, colour = factor(D)), data = dat) +
  geom_point(alpha = 0.5) +
  geom_vline(xintercept = 50, colour = "grey", linetype = 2)+
  stat_smooth(method = "lm", se = F) +
  labs(x = "Test score (X)", y = "Potential Outcome (Y1)")


# simulate the discontinuity
dat <- dat %>%
  mutate(
    y2 = 25 + 40 * D + 1.5 * x + rnorm(n(), 0, 20)
  )

cli::cli_h1("Potential Outcomes after Treatment")
ggplot(aes(x, y2, colour = factor(D)), data = dat) +
  geom_point(alpha = 0.5) +
  geom_vline(xintercept = 50, colour = "grey", linetype = 2) +
  stat_smooth(method = "lm", se = F) +
  labs(x = "Test score (X)", y = "Potential Outcome (Y)")
```




```{r rdd_simulate3, exercise=TRUE, echo=FALSE}

# simultate nonlinearity
dat <- tibble(
    x = rnorm(1000, 100, 50)
  ) %>% 
  mutate(
    x = case_when(x < 0 ~ 0, TRUE ~ x),
    D = case_when(x > 140 ~ 1, TRUE ~ 0),
    x2 = x*x,
    x3 = x*x*x,
    y3 = 10000 + 0 * D - 100 * x + x2 + rnorm(1000, 0, 1000)
  ) %>% 
  filter(x < 280)


# Linear Model for conditional expectation
ggplot(aes(x, y3, colour = factor(D)), data = dat) +
  geom_point(alpha = 0.2) +
  geom_vline(xintercept = 140, colour = "grey", linetype = 2) +
  stat_smooth(method = "lm", se = F) +
  labs(x = "Test score (X)", y = "Potential Outcome (Y)")

# LOESS conditional expectation
ggplot(aes(x, y3, colour = factor(D)), data = dat) +
  geom_point(alpha = 0.2) +
  geom_vline(xintercept = 140, colour = "grey", linetype = 2) +
  stat_smooth(method = "loess", se = F) +
  labs(x = "Test score (X)", y = "Potential Outcome (Y)")


```

#### Questions
- If you estimated the treatment effect using the linear model, would your treatment effect estimate be near the true treatment effect? 



```{r rdd_simulate4, exercise=TRUE, echo=FALSE}
set.seed(12282020)

dat <- tibble(
    x = rnorm(1000, 100, 50)
  ) %>% 
  mutate(
    x = case_when(x < 0 ~ 0, TRUE ~ x),
    D = case_when(x > 140 ~ 1, TRUE ~ 0),
    x2 = x*x,
    x3 = x*x*x,
    # TRUE DGP 
    y3 = 10000 + 0 * D - 100 * x + x2 + rnorm(1000, 0, 1000)
  ) %>% 
  filter(x < 280)

# Fully interacted regression
regression <- lm(y3 ~ D*., data = dat)
  
stargazer(regression, type = "text")

ggplot(aes(x, y3, colour = factor(D)), data = dat) +
  geom_point(alpha = 0.2) +
  geom_vline(xintercept = 140, colour = "grey", linetype = 2) +
  stat_smooth(method = "loess", se = F) +
  labs(x = "Test score (X)", y = "Potential Outcome (Y)")


```

#### Question
- What did you estimate as the treatment effect using a third order polynomial of the running variable? Is it statistically significantly different from zero?
- Does the estimated treatment effect seem correct from the graph?





## The Close Election Design

Lets load the data from Lee et. al. (2004):

```{r}
lmb_data <- read_data("lmb-data.dta") %>% 
  mutate(demvoteshare_c = demvoteshare - 0.5)

```



```{r lmb1, exercise=TRUE, echo=FALSE}

lmb_subset <- lmb_data %>% 
  filter(lagdemvoteshare>.48 & lagdemvoteshare<.52) 
lm_1 <- lm_robust(score ~ lagdemocrat, data = lmb_subset, clusters = id)
lm_2 <- lm_robust(score ~ democrat, data = lmb_subset, clusters = id)
lm_3 <- lm_robust(democrat ~ lagdemocrat, data = lmb_subset, clusters = id)

cli::cli_h1("Original results based on ADA Scores -- Close Elections Sample")
texreg::knitreg(list(lm_1, lm_2, lm_3), type="text")
```


```{r lmb2, exercise=TRUE, echo=FALSE}
#using all data (note data used is lmb_data, not lmb_subset)
lm_1 <- lm_robust(score ~ lagdemocrat, data = lmb_data, clusters = id)
lm_2 <- lm_robust(score ~ democrat, data = lmb_data, clusters = id)
lm_3 <- lm_robust(democrat ~ lagdemocrat, data = lmb_data, clusters = id)

cli::cli_h1("Results based on ADA Scores -- Full Sample")
texreg::knitreg(list(lm_1, lm_2, lm_3), type="text")
```




```{r lmb3, exercise=TRUE, echo=FALSE}
lm_1 <- lm_robust(score ~ lagdemocrat + demvoteshare_c, data = lmb_data, clusters = id)
lm_2 <- lm_robust(score ~ democrat + demvoteshare_c, data = lmb_data, clusters = id)
lm_3 <- lm_robust(democrat ~ lagdemocrat + demvoteshare_c, data = lmb_data, clusters = id)

cli::cli_h1("Results based on ADA Scores -- Full Sample")
texreg::knitreg(list(lm_1, lm_2, lm_3), type="text")
```

```{r lmb4, exercise=TRUE, echo=FALSE}
lm_1 <- lm_robust(score ~ lagdemocrat*demvoteshare_c, 
                  data = lmb_data, clusters = id)
lm_2 <- lm_robust(score ~ democrat*demvoteshare_c, 
                  data = lmb_data, clusters = id)
lm_3 <- lm_robust(democrat ~ lagdemocrat*demvoteshare_c, 
                  data = lmb_data, clusters = id)

cli::cli_h1("Results based on ADA Scores -- Full Sample with linear interactions")
texreg::knitreg(list(lm_1, lm_2, lm_3), type="text")
```


```{r lmb5, exercise=TRUE, echo=FALSE}

lmb_data <- lmb_data %>% 
  mutate(demvoteshare_sq = demvoteshare_c^2)

lm_1 <- lm_robust(score ~ lagdemocrat*demvoteshare_c + lagdemocrat*demvoteshare_sq, 
                  data = lmb_data, clusters = id)
lm_2 <- lm_robust(score ~ democrat*demvoteshare_c + democrat*demvoteshare_sq, 
                  data = lmb_data, clusters = id)
lm_3 <- lm_robust(democrat ~ lagdemocrat*demvoteshare_c + lagdemocrat*demvoteshare_sq, 
                  data = lmb_data, clusters = id)


cli::cli_h1("Results based on ADA Scores -- Full Sample with linear and quadratic interactions")
texreg::knitreg(list(lm_1, lm_2, lm_3), type="text")

```





```{r lmb6, exercise=TRUE, echo=FALSE}

lmb_data <- lmb_data %>% 
  filter(demvoteshare > .45 & demvoteshare < .55) %>%
  mutate(demvoteshare_sq = demvoteshare_c^2)

lm_1 <- lm_robust(score ~ lagdemocrat*demvoteshare_c + lagdemocrat*demvoteshare_sq, 
                  data = lmb_data, clusters = id)
lm_2 <- lm_robust(score ~ democrat*demvoteshare_c + democrat*demvoteshare_sq, 
                  data = lmb_data, clusters = id)
lm_3 <- lm_robust(democrat ~ lagdemocrat*demvoteshare_c + lagdemocrat*demvoteshare_sq, 
                  data = lmb_data, clusters = id)


cli::cli_h1("Results based on ADA Scores -- Close election sample with linear and quadratic interactions")
texreg::knitreg(list(lm_1, lm_2, lm_3), type="text")
```




```{r lmb7, exercise=TRUE, echo=FALSE}
#aggregating the data
categories <- lmb_data$lagdemvoteshare

demmeans <- split(lmb_data$score, cut(lmb_data$lagdemvoteshare, 100)) %>% 
  lapply(mean) %>% 
  unlist()

agg_lmb_data <- data.frame(score = demmeans, lagdemvoteshare = seq(0.01,1, by = 0.01))

#plotting
lmb_data <- lmb_data %>% 
  mutate(gg_group = case_when(lagdemvoteshare > 0.5 ~ 1, TRUE ~ 0))
         
ggplot(lmb_data, aes(lagdemvoteshare, score)) +
  geom_point(aes(x = lagdemvoteshare, y = score), data = agg_lmb_data) +
  stat_smooth(aes(lagdemvoteshare, score, group = gg_group), method = "lm", 
              formula = y ~ x + I(x^2)) +
  xlim(0,1) + ylim(0,100) +
  geom_vline(xintercept = 0.5)

ggplot(lmb_data, aes(lagdemvoteshare, score)) +
  geom_point(aes(x = lagdemvoteshare, y = score), data = agg_lmb_data) +
  stat_smooth(aes(lagdemvoteshare, score, group = gg_group), method = "loess") +
  xlim(0,1) + ylim(0,100) +
  geom_vline(xintercept = 0.5)

ggplot(lmb_data, aes(lagdemvoteshare, score)) +
  geom_point(aes(x = lagdemvoteshare, y = score), data = agg_lmb_data) +
  stat_smooth(aes(lagdemvoteshare, score, group = gg_group), method = "lm") +
  xlim(0,1) + ylim(0,100) +
  geom_vline(xintercept = 0.5)

```




```{r lmb8, exercise=TRUE, echo=FALSE}
smooth_dem0 <- lmb_data %>% 
  filter(democrat == 0) %>% 
  select(score, demvoteshare)
smooth_dem0 <- as_tibble(ksmooth(smooth_dem0$demvoteshare, smooth_dem0$score, 
                                 kernel = "box", bandwidth = 0.1))


smooth_dem1 <- lmb_data %>% 
  filter(democrat == 1) %>% 
  select(score, demvoteshare) %>% 
  na.omit()
smooth_dem1 <- as_tibble(ksmooth(smooth_dem1$demvoteshare, smooth_dem1$score, 
                                 kernel = "box", bandwidth = 0.1))

ggplot() + 
  geom_smooth(aes(x, y), data = smooth_dem0) +
  geom_smooth(aes(x, y), data = smooth_dem1) +
  geom_vline(xintercept = 0.5)
```




```{r lmb9, exercise=TRUE, echo=FALSE}
rdr <- rdrobust(y = lmb_data$score,
                x = lmb_data$demvoteshare, c = 0.5)
summary(rdr)
```




```{r lmb10, exercise=TRUE, echo=FALSE}


DCdensity(lmb_data$demvoteshare, cutpoint = 0.5)

density <- rddensity(lmb_data$demvoteshare, c = 0.5)
rdplotdensity(density, lmb_data$demvoteshare)

```

#### QUESTIONS
- Can you think of another example where you might use the close election design to estimator some average treatment effect?
- To what degree does this study help us understand the importance of incumbency in a Presidential election?  Why/why not?








