---
title: "Causal Inference: <br> *The Mixtape*"
subtitle: "<it>Panel Data</it>"
output: 
  learnr::tutorial:
    css: css/style.css
    highlight: "kate"
runtime: shiny_prerendered
---

## Welcome

This is material for the **Panel Data** chapter in Scott Cunningham's book, [Causal Inference: The Mixtape.](https://mixtape.scunning.com/)

### Packages needed

The first thing you need to do is install a few packages to make sure everything runs:

```{r, eval = FALSE}
install.packages("tidyverse")
install.packages("cli")
install.packages("haven")
install.packages("rmarkdown")
install.packages("learnr")
install.packages("haven")
install.packages("stargazer")

# This Chapter Only
install.packages("plm")
```

### Load

```{r load, warning=FALSE, message=FALSE}
library(learnr)
library(haven)
library(tidyverse)
library(stargazer)

# This Chapter Only
library(plm)

# 10 minute code time limit
options(tutorial.exercise.timelimit = 600)

# read_data function
read_data <- function(df) {
  full_path <- paste0("https://raw.github.com/scunning1975/mixtape/master/", df)
  return(haven::read_dta(full_path))
}

```


## Survey of Adult Service Providers

```{r sasp, exercise=TRUE, echo=FALSE}

sasp <- read_data("sasp_panel.dta")

#-- Delete all NA
sasp <- na.omit(sasp)

#-- order by id and session 
sasp <- sasp %>% 
  arrange(id, session)

#Balance Data
balanced_sasp <- make.pbalanced(sasp, balance.type = "shared.individuals")

#Demean Data
balanced_sasp <- balanced_sasp %>% 
  mutate( 
    demean_lnw = lnw - ave(lnw, id),
    demean_age = age - ave(age, id),
    demean_asq = asq - ave(asq, id),
    demean_bmi = bmi - ave(bmi, id),
    demean_hispanic = hispanic - ave(hispanic, id),
    demean_black = black - ave(black, id),
    demean_other = other - ave(other, id),
    demean_asian = asian - ave(asian, id),
    demean_schooling = schooling - ave(schooling, id),
    demean_cohab = cohab - ave(cohab, id),
    demean_married = married - ave(married, id),
    demean_divorced = divorced - ave(divorced, id),
    demean_separated = separated - ave(separated, id),
    demean_age_cl = age_cl - ave(age_cl, id),
    demean_unsafe = unsafe - ave(unsafe, id),
    demean_llength = llength - ave(llength, id),
    demean_reg = reg - ave(reg, id),
    demean_asq_cl = asq_cl - ave(asq_cl, id),
    demean_appearance_cl = appearance_cl - ave(appearance_cl, id),
    demean_provider_second = provider_second - ave(provider_second, id),
    demean_asian_cl = asian_cl - ave(asian_cl, id),
    demean_black_cl = black_cl - ave(black_cl, id),
    demean_hispanic_cl = hispanic_cl - ave(hispanic_cl, id),
    demean_othrace_cl = othrace_cl - ave(lnw, id),
    demean_hot = hot - ave(hot, id),
    demean_massage_cl = massage_cl - ave(massage_cl, id)
  )

#-- POLS
cli::cli_h1("Pooled OLS") 
ols <- lm_robust(lnw ~ age + asq + bmi + hispanic + black + other + asian + schooling + cohab + married + divorced + separated + 
           age_cl + unsafe + llength + reg + asq_cl + appearance_cl + provider_second + asian_cl + black_cl + hispanic_cl + 
           othrace_cl + hot + massage_cl, data = balanced_sasp)
summary(ols)


#-- FE
cli::cli_h1("Fixed Effects") 
formula <- as.formula("lnw ~ age + asq + bmi + hispanic + black + other + asian + schooling + 
                      cohab + married + divorced + separated + 
                      age_cl + unsafe + llength + reg + asq_cl + appearance_cl + 
                      provider_second + asian_cl + black_cl + hispanic_cl + 
                      othrace_cl + hot + massage_cl")

model_fe <- lm_robust(formula = formula,
                  data = balanced_sasp, 
                  fixed_effect = ~id, 
                  se_type = "stata")

summary(model_fe)

#-- Demean OLS
cli::cli_h1("Demeaned OLS") 
dm_formula <- as.formula("demean_lnw ~ demean_age + demean_asq + demean_bmi + 
                demean_hispanic + demean_black + demean_other +
                demean_asian + demean_schooling + demean_cohab + 
                demean_married + demean_divorced + demean_separated +
                demean_age_cl + demean_unsafe + demean_llength + demean_reg + 
                demean_asq_cl + demean_appearance_cl + 
                demean_provider_second + demean_asian_cl + demean_black_cl + 
                demean_hispanic_cl + demean_othrace_cl +
                demean_hot + demean_massage_cl")

ols_demean <- lm_robust(formula = dm_formula, 
                data = balanced_sasp, clusters = id,
                se_type = "stata")

summary(ols_demean)

```

#### QUESTIONS
- Interpret the effect of natural log of session length on the natural log of hourly wage.  Describe the economic theory that might explain this relationship?  (HINT: Consider the role that supplier fixed versus variable costs may have on the hourly wage.)
- Becker described discrimination in terms of ``taste based``.  This meant that social interactions with people of the other race were factors into marginal cost.  Given that these persist, what does this imply about the effect that competition is having on discrimination?
- Hamermesh and Biddle suggest that beauty is valued on the market.  Describe some reasons why there is no effect on client beauty once we use the within estimators?
- What other interesting results did you find in this analysis?  Which ones surprised you and which ones were intuitive and why?

